---
title: "Basir"
---


```{r}
library(tidyverse)
# import Data
History_of_Baseball <- read_csv("data/The History of Baseball dataset/hall_of_fame.csv")

str(History_of_Baseball)
head(History_of_Baseball)
```

----
2.3 Pitching table

playerID       Player ID code
yearID         Year
stint          player's stint (order of appearances within a season)
teamID         Team
lgID           League
W              Wins
L              Losses
G              Games
GS             Games Started
CG             Complete Games 
SHO            Shutouts
SV             Saves
IPOuts         Outs Pitched (innings pitched x 3)
H              Hits
ER             Earned Runs
HR             Homeruns
BB             Walks
SO             Strikeouts
BAOpp          Opponent's Batting Average
ERA            Earned Run Average
IBB            Intentional Walks
WP             Wild Pitches
HBP            Batters Hit By Pitch
BK             Balks
BFP            Batters faced by Pitcher
GF             Games Finished
R              Runs Allowed
SH             Sacrifices by opposing batters
SF             Sacrifice flies by opposing batters
GIDP           Grounded into double plays by opposing batter
```{r}
History_of_Baseball <- read_csv("data/The History of Baseball dataset/pitching.csv")

str(History_of_Baseball)
head(History_of_Baseball)
```

```{r}
ggplot(History_of_Baseball) +
  geom_line(aes(x=year, y=w))

```

