---
title: "Report Draft"
format: html
editor: visual
runtime: shiny
---

```{r setup, include=FALSE}
library(shiny)
library(tidyverse)
library(plotly)
library(shinylive)


pitching <- read.csv("data/The History of Baseball dataset/pitching.csv")
batting <- read.csv("data/The History of Baseball dataset/batting.csv")

# Group years into decades
pitching <- pitching %>%
  mutate(decade = paste0(floor(year / 10) * 10, "s"))

# Add batting average to pitching
bat_sm <- batting %>% 
  mutate(batt_avg = h/ab) %>% 
  select(player_id, year, batt_avg)

pitching_2 <- pitching %>% 
  filter(year >= 1985) %>% 
  left_join(bat_sm)
```

### Introduction

Baseball is one of the most popular sports in the US with millions of viewers each year. There are many factors that contribute to team wins and we would like to explore potential factors contributing to these wins. With this popularity comes a plethora of interesting data, from batting averages to different measures of success in the field to All-Star players per team. Using this data going as far back as the 1870s, we decided to look into what factors can predict team wins over time. With billions of dollars pouring into the MLB (Major League Baseball) each year, finding potential predictors of which teams will win is quite literally a billion dollar question. Throughout this report, we will explore different relationships between the variable `wins` and several predictor variables.

In order to best understand this report, a few terms must be clarified. Firstly, `wins` refer to the number of games won by a team, while `losses` refers to the number of games lost by a team. A strikeout is when the pitcher throws three strikes, valid pitches that the batter does not hit, causing the batter to be out and finished with their batting turn. Finally, batting average is a decimal value representing the number of successful hits made by a single batter divided by their total number of hitting attempts.

This data was acquired from Kaggle. This data set, titled 'The History of Baseball' is a reformatted version of a famous data set by Sean Lahman. This data, originally published in the 1990s was the first of its kind, making raw baseball data accessible to the public. This data was most recently updated for the 2015 season, and contains data between 2015 and 1871. Lahman's data set is actually a collection of several data sets, ranging from pitching data, to data on team managers. For our analysis, we focus on the pitching and batting data sets, which contain information on wins, losses, pitching stats, batting stats, and salary information per player per year. In order to complete our analysis, we have joined the batting and pitching data sets together, added a new variable `batting_avg`, and cleaned the variable names. The variables are a mix between categorical (player names, team names, etc) and quantitative (wins, losses, batting average, etc).


```{r}
#TODO: Add numerical summaries?
```

## Wins and Losses per Team
```{r}
# TODO: Add alt-cap, make sure color palette is accessible

# Define UI for selecting a decade
selectInput("selected_decade", 
            "Choose a Decade:", 
            choices = unique(pitching$decade), 
            selected = unique(pitching$decade)[1])

# Render the plot dynamically
renderPlot({
  filtered_data <- pitching %>% filter(decade == input$selected_decade)
  
  ggplot(filtered_data, aes(x = team_id, group = league_id)) +
    geom_bar(aes(y = w, fill = "Wins"), stat = "identity", position = "dodge",alpha = 0.4) +
    geom_bar(aes(y = l, fill = "Losses"), stat = "identity", position = "dodge",alpha = 0.4) +
    labs(title = paste("Wins and Losses in", input$selected_decade), 
         x = "Year", y = "Count") +
    scale_fill_manual(values = c("Wins" = "blue", "Losses" = "red")) +
    theme_minimal()+
    theme(axis.text.x = element_text(angle = 45, hjust = 1))# Rotate team labels for clarity

})

```

### Data Insights

What are the important takeaways from the data? What was interesting?

Why do these takeaways matter?

Was there anything surprising?

Overall, what do you want the audience to walk away with? What do you want them to understand about your data and research 
questions?!?


## Strikeouts VS Walks
```{r}
# TODO: Add alt-cap, make sure color palette is accessible

# UI to select filters
sidebarLayout(
  sidebarPanel(
    sliderInput("top_n", 
                "Select Top N Players by Strikeouts:", 
                min = 5, max = 50, value = 10, step = 5),
    selectInput("filter_team", 
                "Filter by Team (Optional):", 
                choices = c("All", unique(pitching$team_id)), 
                selected = "All")
  ),
  mainPanel(
    plotOutput("hr_bb_plot")
  )
)

# Server logic
renderPlot({
  # Filter and sort data
  filtered_data <- pitching %>%
    filter(if (input$filter_team != "All") team_id == input$filter_team else TRUE) %>%
    arrange(desc(hr)) %>%
    slice_head(n = input$top_n)  # Top N players
  
  ggplot(filtered_data, aes(x = bb, y = so, color = player_id, label = player_id)) +
    geom_point(size = 3, alpha = 0.7) +
    geom_text(vjust = -0.5, size = 3) +
    labs(title = "Strikeouts vs. Walks", 
         x = "Walks (BB)", y = "Strikeouts (SO)") +
    theme_minimal() +
    theme(legend.position = "none")

})
```

### Data Insights

What are the important takeaways from the data? What was interesting?

Why do these takeaways matter?

Was there anything surprising?

Overall, what do you want the audience to walk away with? What do you want them to understand about your data and research 
questions?!?


## Strikeouts VS Wins Since 1985
```{r}
# still a WIP, feel free to mess around with it
# TODO: Add alt-cap, make sure color palette is accessible

pitching %>% 
  filter(year >= 1985, so >= 5) %>% 
  ggplot(aes(x = so, y = w)) +
  geom_point() +
  geom_smooth(method = "lm") +
  labs(title = "Strikeouts vs Wins since 1985", x = "Strikeouts (SO)", y = "Wins (W)")
```

### Data Insights

What are the important takeaways from the data? What was interesting?

Why do these takeaways matter?

Was there anything surprising?

Overall, what do you want the audience to walk away with? What do you want them to understand about your data and research 
questions?!?


## Batting Average VS Wins Since 1985
```{r}
# still a WIP, feel free to mess around with it maybe add an interactive team selector?
# TODO: Add alt-cap, make sure color palette is accessible

bat_sm <- batting %>% 
  mutate(batt_avg = h/ab) %>% 
  select(player_id, year, batt_avg)

pitching_2 <- pitching %>% 
  filter(year >= 1985) %>% 
  left_join(bat_sm)
  
pitching_2 %>% 
  filter(!is.na(batt_avg)) %>% 
  ggplot(aes(x = batt_avg, y = w, group = team_id)) +
  geom_boxplot()
```

### Data Insights

What are the important takeaways from the data? What was interesting?

Why do these takeaways matter?

Was there anything surprising?

Overall, what do you want the audience to walk away with? What do you want them to understand about your data and research 
questions?!?



### Conclusion/Big Picture

How do the insights connect to answer your research question?

What improvements might someone make to your analysis?

Are there any limitations or weaknesses of your data / analysis?
